---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pslr

<!-- badges: start -->
[![Verify](https://github.com/bart-turczynski/pslr/actions/workflows/verify.yml/badge.svg)](https://github.com/bart-turczynski/pslr/actions/workflows/verify.yml)
[![CRAN status](https://www.r-pkg.org/badges/version/pslr)](https://CRAN.R-project.org/package=pslr)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/pslr)](https://CRAN.R-project.org/package=pslr)
[![Codecov coverage](https://codecov.io/gh/bart-turczynski/pslr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/bart-turczynski/pslr)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

A focused, spec-complete implementation of the
[Public Suffix List](https://publicsuffix.org) (PSL) for R. `pslr` bundles a
reproducible, pinned PSL snapshot and implements the official prevailing-rule
algorithm to answer public-suffix (eTLD) and registrable-domain (eTLD+1)
queries.

- Distinguishes the **ICANN** and **PRIVATE** rule sections.
- Accepts Unicode, ASCII, and A-label hostnames via `punycoder`
  canonicalization; returns ASCII or Unicode output.
- Works fully **offline** from the bundled snapshot; an explicit, validated
  `psl_refresh()` is the only network path.
- Matcher compiled with `cpp11`; **no external system library** required.

## Installation

Install the released version from CRAN:

```{r eval=FALSE}
install.packages("pslr")
```

Or the development version from GitHub:

```{r eval=FALSE}
# install.packages("pak")
pak::pak("bart-turczynski/pslr")
```

`pslr` depends on [`punycoder`](https://cran.r-project.org/package=punycoder),
which is installed automatically from CRAN.

## Usage

```{r usage}
library(pslr)

public_suffix("www.example.co.uk")
registrable_domain("www.example.co.uk")

# ICANN vs PRIVATE sections
public_suffix("user.github.io")
public_suffix("user.github.io", section = "icann")

# Explicit membership vs the implicit default rule
is_public_suffix("madeuptld")                 # implicit "*"
is_public_suffix("madeuptld", unknown = "na") # explicit only

# Split a host, or inspect the prevailing rule
suffix_extract("blog.user.github.io")
public_suffix_rule("a.b.kobe.jp")
```

See `vignette("introduction", package = "pslr")` for the full tour: section
choice, the unknown-suffix policy, IDN output, terminal dots, refresh and
activation, reproducibility, and security notes.

## Reproducibility

A result depends on both which list answered and how hosts were normalized.
`psl_version()` reports the active-list provenance plus the runtime
normalization identifiers; record it alongside reproducibility-sensitive output.

## How pslr compares to other PSL libraries

The [Public Suffix List website](https://publicsuffix.org/learn/) catalogs
implementations in C, C#, C++, Go, Haskell, Java, JavaScript, Perl, PHP, Python,
Ruby, Rust, Swift, and more — but no R. `pslr` fills that gap, and it is built
to be a *reproducibility- and correctness-first* engine rather than a quick
suffix splitter.

The table compares `pslr` with a representative set of the most established
libraries from that catalog, across the dimensions that matter for correct,
auditable suffix handling. ✅ first-class · ◐ partial/limited · ❌ absent.

| Library (language) | Full algorithm (`*`/`!`) | ICANN / PRIVATE / both | IDN + Punycode | Offline default + explicit refresh | Queryable provenance | Compiled core | Strict input validation | Unlisted-TLD policy configurable |
|---|:--:|:--:|:--:|:--:|:--:|:--:|:--:|:--:|
| **pslr (R)** | ✅ | ✅ 3-way | ✅ in & out | ✅ bundled; validated HTTPS-only `psl_refresh()` | ✅ list **and** normalization identity | ✅ `cpp11` | ✅ rejects URLs, IPv4, IPv6 | ✅ `unknown=` |
| libpsl (C) | ✅ | ✅ 3-way | ✅ | ✅ bundled; OS-package update | ◐ sha1 + mtime | ✅ | ◐ | ✅ |
| x/net/publicsuffix (Go) | ✅ | ◐ `icann` flag | ❌ ASCII only | ✅ embedded; bump module | ◐ date constant | ✅ | ◐ | ❌ |
| publicsuffix-go (Go) | ✅ | ◐ toggle private | ✅ | ✅ embedded | ❌ | ❌ | ◐ | ✅ |
| psl crate (Rust) | ✅ | ◐ per-suffix type | ✅ | ✅ compiled-in | ◐ dated releases | ✅ | ❌ | ◐ |
| tldextract (Python) | ✅ | ◐ private on/off | ✅ | ⚠️ network-first; offline opt-in | ❌ | ❌ | ◐ | ❌ |
| publicsuffixlist (Python) | ✅ | ◐ exclude private | ✅ | ✅ bundled + updater | ◐ date in version | ❌ | ❌ | ❌ |
| public_suffix (Ruby) | ✅ | ◐ `ignore_private` | ❌ caller pre-encodes | ✅ bundled | ❌ | ❌ | ◐ | ✅ |
| php-domain-parser (PHP) | ✅ | ✅ 3-way | ✅ | ⚠️ not bundled; PSR-16 cache | ◐ `isKnown` flags | ❌ | ◐ | ◐ |
| tldts (JS/TS) | ✅ | ◐ `allowPrivateDomains` | ✅ | ✅ embedded; npm bump | ◐ submodule pin | ◐ optional WASM | ✅ | ◐ |
| Guava `InternetDomainName` (Java) | ◐ | ✅ registry vs public | ✅ | ◐ in-jar; bump Guava | ❌ | ❌ | ◐ | ❌ |

### What `pslr` does differently

- **Provenance is best-in-class.** `psl_version()` records the list identity
  (source, commit, date, SHA-256) *and* the normalization identity (normalizer
  package + version, profile, Unicode version). A PSL answer depends on both
  *which list* answered and *how the host was normalized* — `pslr` is the only
  surveyed library that surfaces both, so a result is genuinely reproducible.
- **Conformance is verified.** The package ships the official upstream
  `tests.txt` vectors, pinned in lockstep with the bundled snapshot and run on
  every check.
- **Offline by default, one hardened network path.** Every query works with zero
  network access; `psl_refresh()` is the *only* code that touches the network —
  HTTPS-only, credential- and downgrade-rejecting, size-capped, atomically
  committed. This is stricter than network-first designs and pairs with the
  reproducibility story rather than fighting it.
- **Strict, policy-driven validation.** URLs, IPv6, dotted-decimal IPv4, and
  malformed labels are rejected; `invalid = "na"` / `"error"` lets you choose
  silent `NA` or a hard stop. Many libraries are deliberately lenient.
- **Vectorized, `NA`-safe, name-preserving.** Every function operates on a whole
  character vector — the idiomatic shape for data work in R.

### Trade-offs

- **Session-global active list.** There is no per-call list switching yet; the
  active list is per-session state (`psl_use()` / `psl_refresh()`).
- **Hostnames, not URLs.** URL-shaped input is rejected by design; parse the host
  out first or use [`rurl`](https://bart-turczynski.github.io/rurl/).
- **No network-first auto-fetch.** Refreshing is always explicit — a deliberate
  choice for reproducibility, not a convenience feature.

## Development

Install dependencies plus the dev tooling used by the checks:

```sh
Rscript -e 'pak::local_install_deps(dependencies = TRUE)'
```

Run the same verification CI runs (lint + `R CMD check --as-cran`):

```sh
Rscript -e 'lints <- lintr::lint_package(); if (length(lints)) { print(lints); quit(status = 1) }' && Rscript -e 'rcmdcheck::rcmdcheck(args = "--as-cran", error_on = "warning")'
```

`R CMD check` runs the testthat and cucumber specs, so the behaviour specs are
verified as part of the check. A non-CRAN performance benchmark and its release
gate live in [`bench/benchmark.R`](https://github.com/bart-turczynski/pslr/blob/main/bench/benchmark.R); recorded reference
results are in [`docs/benchmarks.md`](https://github.com/bart-turczynski/pslr/blob/main/docs/benchmarks.md).

### Project layout

- `R/` — package source (edit roxygen comments here, not `man/` or `NAMESPACE`).
- `src/` — the `cpp11` matcher core.
- `man/` — generated help pages (`devtools::document()`).
- `tests/testthat/` — testthat tests and cucumber feature specs.
- `vignettes/` — long-form documentation.
- `data-raw/` — the deterministic snapshot regeneration pipeline.
- `docs/` — durable project context (`PRD.md`, `architecture.md`, `benchmarks.md`).

## Related packages

`pslr` is part of a small ecosystem of R packages by the same author:

- **[punycoder](https://bart-turczynski.github.io/punycoder/)** — the Punycode and IDNA codec that `pslr` uses for host canonicalization before PSL matching. Use it directly for raw Unicode ↔ ACE round-trips.
- **[rurl](https://bart-turczynski.github.io/rurl/)** — full URL parsing, normalization, cleaning, and joining toolkit. Uses `pslr` as its PSL engine; reach for it when you need more than domain extraction.

## License

Package code is MIT licensed. The bundled Public Suffix List data
(`inst/extdata/`) is distributed under the Mozilla Public License 2.0; see
`inst/NOTICE` and `inst/extdata/PSL-LICENSE`.
